---
title: "DanielsAms394Hw1"
author: "William Daniels"
date: "2024-05-31"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 1

Authored by William Daniels, on May 31, 2024 for AMS 392: Statistical laboratory

## Part 1

### Problem 1

Consider the following heights: 1.45, 2.22, 1.65, 1.25, 1.58, 1.87, 1.93, 1.42, 1.26, 1.57, 1.52, 1.95. Use R script to answer the following questions.

-   Assign all these heights as a vector height

-   Compute the mean and standard deviation of height

-   What is the length of "height?"

-   How many heights are less than 1.65?

-   Show if each height is larger than 1.60 and smaller than 1.75.

#### Code

```{r}
height <- c(1.45, 2.22, 1.65, 1.25, 1.58, 1.87, 1.93, 1.42, 1.26, 1.57, 1.52, 1.95)

mean(height)
sd(height)

length(height)

sum(height<1.65)

height>1.6 & height<1.75
sum(height>1.6 & height<1.75)
```

#### Solutions

-   The mean height is 1.639 and the standard deviation is 0.297

-   There are 12 heights

-   7 heights are less than 1.65

-   There is 1 height between 1.6 and 1.75, and it is the third height in the list

### Problem 2

Using the following script, we can generate a 3x4 matrix: `tmp <- matrix(rnorm(12), 3, 4)`. Answer the following:

-   Compute the sum of the second and third rows (The expected result should be the sum of each column). (by sum function)

-   Compute the product of second and fourth columns. (The expected result should be the product of each column). (by prod function)

-   Show the dimension of the matrix.

-   Use ”cat” function to output elements in the second row that are less than 0.2. (use the help document to check the cat function)

#### Code

```{r}
m <- matrix(rnorm(12), 3, 4)
m

sum(m[2,])
sum(m[3,])

prod(m[,2])
prod(m[,4])

dim(m)

cat(m[2,][m[2,]<0.2], sep=", ")
```

#### Solutions

-   The sum of the second row is -0.169 and the sum of the third row is 4.592

-   The product of the second column is -0.521 and the product of the fourth column is 0.111

-   The dimension on the matrix is 3, 4

-   Elements in the second row less than 0.2 are: 0.091, -0.598, and 0.107

## Part 2

### Problem 1

-   Calculate the probability of $x^2 < 8$, if $x$ follows a normal distribution with mean $1$ and standard deviation $2$.

-   Calculate the probability of getting 3 out of 8 successes in a binomial distribution with probability 0.5

-   Calculate the probability of $x < 2$, if $x$ follows a chi-square distribution with degree of freedom 7

-   Calculate the probability of $5 < x < 7$, if $x$ follows an exponential distribution with rate 4

-   Calculate the probability of $x=7$, if $x$ follows a poisson distribution with mean 5

#### Code

```{r}
trials <- 1000000
sum(rnorm(trials, 1, 2)^2 < 8) / trials

dbinom(3, 8, 0.5)

pchisq(2, 7)

pexp(7, 4) - pexp(5, 4)

dpois(7, 5)
```

#### Solutions

-   0.792

-   0.219

-   0.040

-   0.000

-   0.104

### Problem 2

-   Check the normality of the variable "std.wright" in "wright" data set in package "ISwR" with QQ plot or histogram

-   Test if the mean of "std.wright" is greater than 400 with significance level $\alpha = 0.05$

#### Code

```{r}
#install.packages("ISwR")
library("ISwR")
qqnorm(wright$std.wright)
qqline(wright$std.wright)

t.test(wright$std.wright, alternative="greater", mu=400, conf.level=0.95)
```

#### Solutions

-   The QQ plot indicates that the bulk of the data is roughly normal, but the "tails" of the distribution are likely "heavier"
-   If we allow the assumption that the data is normal, we can use a t-test since the standard deviation of the population is unknown. Since the 0.95 confidence interval exceeds 400, we conclude that the true mean is greater than 400

## Part 3

### Problem 1

Calculate the probability of each of the following events:

-   A normally distributed variable with mean 10 and standard deviation 3 is less than 10

-   $X<4$ in a chi-square distribution with 6 degrees of freedom

-   Getting 4 out of 6 successes in a binomial distribution with probability 0.85

-   $X=5$ in a Poisson distribution with $\lambda$ is 10

#### Code

```{r}
pnorm(10, mean=10, sd=3)

pchisq(4, 6)

dbinom(4, 6, 0.85)

dpois(5, 10)
```

#### Solutions

-   0.500

-   0.323

-   0.176

-   0.038

### Problem 2

Generate 1,000 exponentially distributed random variables with rate 5, and plot their empirical distribution function

#### Code

```{r}
samples <- rexp(1000, 5)
plot(ecdf(samples), main="Empriric distribution of exponential samples")
```

#### Solution

See above plot.

### Exercise 0

Generate 1000 samples from a normal distribution with mean 2 and standard deviation 3, and approximate the 25th, 50th, and 75th percentile.

#### Code

```{r}
samples <- rnorm(1000, 2, 3)
boxplot(samples)
bpstats <- boxplot.stats(samples)
text(x = 1.36, labels = bpstats$stats, y = bpstats$stats)

quantile(samples, probs = c(0.25, 0.5, 0.75))
```

#### Solutions

Excluding outliers, $Q_1 = -0.034$, $Q_2 = 1.994$, and $Q_3 = 4.087$. Including outliers, $Q_1 = -0.031$, $Q_2 = 1.994$, and $Q_3 = 4.087$
